Deconvolution to FANS4

### 2020-08-26 deconvolution/apply_dtangle_FANS4.Rmd
add upload

knit2synapse::createAndKnitToFolderEntity(file = "deconvolution/apply_dtangle_FANS4.Rmd",
                                 parentId ="syn22321035",
                                 folderName = "Deconvolution results",
                                 knitmd=FALSE)
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
library(data.table)
library(dtangle)
library(edgeR)
library(variancePartition)
library(BiocParallel)

library(synapser)
library(future)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

source("../common_functions/common_functions.R")

@@

```{r read.data}

# Read reference panel
######################
geneExpr_FANS4_lcpm = read.csv(synGet('syn22321183')$path, header=TRUE)
rownames(geneExpr_FANS4_lcpm) = geneExpr_FANS4_lcpm[,1]
geneExpr_FANS4_lcpm = geneExpr_FANS4_lcpm[,-1]
metadata_FANS4 = read.csv(synGet("syn22321189")$path, header=TRUE, row.names=1)

# read Raw counts
#################
file_counts_synid = c('syn21888969', 'syn21862455', 'syn21887889', 'syn21825508')

@@

cellFracEstimates2 = dtangle2( Y = t(geneExpr_logCPM), reference = t(geneExpr_FANS4_lcpm[idx,]))

cor(cellFracEstimates$estimates, cellFracEstimates2$estimates, method="spearman")
```

### Data preprocessing
```{r preprocess.data}
# Get genotype ids from synapse
GENOTYPE_ID = 'syn16816490'
ALL_USED_IDs = c(ALL_USED_IDs, GENOTYPE_ID)
GENOTYPE = downloadFile(GENOTYPE_ID) %>% 
  dplyr::select(Individual_ID, `SNP_report:Genotyping_Sample_ID`, `SNP_report:Exclude?`) %>% 
  dplyr::inner_join(ANCESTRY) %>% 
  dplyr::filter(is.na(`SNP_report:Exclude?`))

# Merge metadata
METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>% 
  dplyr::rename(Region = `rnaSeq_dissection:Brain_Region`,
                PMI = `PMI_(in_hours)`,
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                GenotypeExclude = `SNP_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Cohort, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
  dplyr::filter(Cohort %in% "NIMH-HBCC") %>% 
  dplyr::filter(SampleID %in% colnames(COUNT), !is.na(SampleID)) %>% 
  dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "BP", Other = "undetermined", Control = "Control", SCZ = "SCZ"))

ind = METADATA$SampleID[which(METADATA$ReportExclude == 1 | METADATA$GenotypeExclude)]
writeLines(paste('Following',length(ind),'samples are marked exclude'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Ethnicity) | is.na(METADATA$Institution) | is.na(METADATA$Dx)]
writeLines(paste('Following', length(ind), 'counts are missing any metadata'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$PMI)]
writeLines(paste('Following', length(ind), 'counts are missing PMI'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Reported_Gender)]
writeLines(paste('Following', length(ind), 'counts are missing gender'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$ageOfDeath)]
writeLines(paste('Following', length(ind), 'counts are missing age of death'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$EV.1)]
writeLines(paste('Following', length(ind), 'counts are missing ancestry information'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))
```

```{r preprocess.data1, results='asis'}
# Match covariates to expression data
indToRetain = intersect(METADATA$SampleID, colnames(COUNT))

rownames(COUNT) <- COUNT$Geneid
COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$SampleID
METADATA = METADATA[indToRetain,]

METADATA %>% 
  group_by(Dx) %>% 
  dplyr::summarise(count = n()) %>% 
  kable()

FactorCovariates <- c('Individual_ID', "Reported_Gender", "LibraryBatch", "Dx", "FlowcellBatch", "RibozeroBatch")
ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", "AlignmentRate",
                    "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){
  x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], function(x){
  x = as.numeric(as.character(gsub('[\\,\\%]','',x)))
})

```{r store, eval=FALSE}
# Code
CODE <- Folder(name = "Meta-analysis on sex", parentId = "syn22315403") # Make sure these match the folder name and parent Id in the first chunk.
CODE <- synStore(CODE)

# Store meta-analysis
file = "cellFracEstimates.RDS"
saveRDS(cellFracEstimates, file=file)
nEXP_OBJ = File(file, 
                name = 'Result of dtangle', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = Ids, 
                    executed = thisFile)
res_acc %>%
  tibble::rownames_to_column(var = "gene_id") %>% 
  write.table(file = 'CMC_acc_Sex_meta_analysis.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_acc_Sex_meta_analysis.tsv', 
                name = 'Sex Normalised, meta-analysis (ACC)', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = Ids, 
                    executed = thisFile)
```

>>>>

knit2synapse::createAndKnitToFolderEntity(file = "deconvolution/apply_dtangle_FANS4.Rmd",
                                 parentId ="syn22321035",
                                 folderName = "Deconvolution results",
                                 knitmd=TRUE)
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
suppressPackageStartupMessages({
library(data.table)
library(dtangle)
library(edgeR)
library(variancePartition)
library(BiocParallel)
library(RNOmni)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)
library(synapser)
library(future)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')
})
source("../common_functions/common_functions.R")

@@

```{r read.data}
# Read reference panel
######################
geneExpr_FANS4_lcpm = read.csv(synGet('syn22321183')$path, header=TRUE)
rownames(geneExpr_FANS4_lcpm) = geneExpr_FANS4_lcpm[,1]
geneExpr_FANS4_lcpm = geneExpr_FANS4_lcpm[,-1]
metadata_FANS4 = read.csv(synGet("syn22321189")$path, header=TRUE, row.names=1)

# read Raw counts
#################
file_counts_synid = c('syn21888969', 'syn21862455', 'syn21887889', 'syn21825508')
ALL_USED_IDs = c('syn22321183', file_counts_synid)

@@

# cellFracEstimates2 = dtangle2( Y = t(geneExpr_logCPM), reference = t(geneExpr_FANS4_lcpm[idx,]))

# cor(cellFracEstimates$estimates, cellFracEstimates2$estimates, method="spearman")
```

### Data preprocessing
```{r preprocess.data}
synId_covariates = c('syn21862454', 'syn21887888', 'syn21825507') # 'syn21888968'

COV_list = lapply(synId_covariates, function(id){
	df = read.table( synGet(id)$path, header=TRUE, stringsAsFactors=FALSE)
	df$SampleID = as.character(df$SampleID)
	if( any(colnames(df) == "Dx.Tissue") ){
		df$Dx = gsub("^(\\S+)_(\\S+)$", "\\1", df$Dx.Tissue)
		df$Tissue = gsub("^(\\S+)_(\\S+)$", "\\2", df$Dx.Tissue)
	}	
	if( !any("Institution" == colnames(df)) ){
		df$Institution = "HBCC"

	}
	if( id == 'syn21862454' ){
		df$Tissue = "DLPFC"
	}
	if( id == 'syn21887888' ){
		df$Tissue = "sgACC"
	}
	if( !any('Tissue' == colnames(df))){
		stop(id)
	}
	df
	})

COVARIATES = gtools::smartbind( list = COV_list)

df = with(COVARIATES, data.frame(SampleID, Study = paste(Institution, Tissue)))
df$Study = gsub("Penn|Pitt", "MSSM", df$Study)

ids = unique(df$Study)
sampleByStudy = lapply( ids, function(stdy){
	df[df$Study == stdy, "SampleID"]
	})
names(sampleByStudy) = ids

# only keep individuals older than 17
COVARIATES = COVARIATES[COVARIATES$ageOfDeath > 17,]

```{r deconv.compare.covariates}

deconvEstimates = data.frame(cellFracEstimates$estimates)

f = log
# f = rankNorm

df_deconv_test = lapply( names(sampleByStudy) , function(stdy){

	smplids = sampleByStudy[[stdy]]

	deconvEst = deconvEstimates[smplids,]

	df = merge(deconvEst, COVARIATES[COVARIATES$SampleID %in% smplids,], by.x="row.names", by.y="SampleID")

	form = cbind(GABA 	= f(GABA), 
				GLU 	= f(GLU), 
				Olig 	= f(Olig),
				MgAs 	= f(MgAs)) ~ ageOfDeath + Reported_Gender + PMI + RIN + RIN2 + Dx 
	fit = lm(form, df)

	res = lapply( summary(fit), function(x) data.frame(p = -log10(coef(x)[-1,4]) * sign(coef(x)[-1,1]) ))

	df_return = do.call(cbind, res)
	colnames(df_return) = gsub("Response ", "", names(res))
	data.frame(Study = stdy, coef = rownames(df_return), df_return)

	})

df_deconv_test = do.call(rbind, df_deconv_test)

df_deconv_test_melt = reshape2::melt(df_deconv_test, id.vars=c("Study", "coef"))
n_tests = nrow(df_deconv_test_melt)
df_deconv_test_melt$isSignif = with(df_deconv_test_melt, abs(value) > -log10(0.05 / n_tests))

lim = max(abs(df_deconv_test_melt$value))

ratio = with(df_deconv_test_melt, length(unique(coef)) / length(unique(Study)))

ggplot(df_deconv_test_melt, aes(Study, coef, fill=value, label=ifelse(isSignif, '*', ''))) + geom_tile() + theme_bw() + theme(aspect.ratio = ratio, panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.x=element_text(angle=45, hjust=1)) + scale_fill_gradient2(high = "red", mid = "white", low="blue", limits=c(-lim, lim), name="-log10 p") + facet_wrap( ~ variable) + geom_text()
```

```{r combine.samples}

df = merge(deconvEstimates, COVARIATES, by.x="row.names", by.y="SampleID")

form = ~ (1|Individual_ID) + ageOfDeath + (1|Reported_Gender) + PMI + RIN + (1|Dx) + (1|Tissue)
vp = fitExtractVarPartModel( t(log(df[,colnames(deconvEstimates)])), form, df, BPPARAM=SerialParam())

plotPercentBars( vp )
```

```{r cca, fig.width=5, fig.height=5}
form = ~ Individual_ID + ageOfDeath + Reported_Gender + PMI + RIN + Dx + Tissue
C = canCorPairs(form, df)
plotCorrMatrix(C)
```

```{r store, eval=TRUE}
# Code
CODE <- Folder(name = "Deconvolution results", parentId = "syn22321035") # Make sure these match the folder name and parent Id in the first chunk.
CODE <- synStore(CODE)

# Store meta-analysis
file = "cellFracEstimates_dtangle_FANS4.tsv"
# saveRDS(data.frame(cellFracEstimates$estimates), file=file)
write.table(cellFracEstimates$estimates, quote=FALSE, sep='\t', file=file)

nEXP_OBJ = File(file, 
                name = 'Estimates from dtangle', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = Ids, 
                    executed = thisFile)
```
